---
title: "2_EDA"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
library(ggplot2)
setwd(here::here("code"))
source("1_data_prep.R")
```

# TODO

-   data = filtered by 2+ surveys, from the main 7 watersheds, only new egg masses
-   *delete anything that is filtering/selecting 7 watersheds in this file*

# data manipulation

tables that are helpful for creating many graphs

```{r summary_new_eggs}
#summary information of egg masses count

#total number of egg masses is 3939
sum(data$NumberofEggMasses, na.rm = TRUE)

#maximum number of egg masses for a given survey is 25, while the mean is 0.857
summary(data$NumberofEggMasses,na.rm = TRUE)
```

#### dataframe: "statistics"

1.  number of survey
2.  mean number of new egg masses per survey
3.  total number of egg masses per watershed per site per year
4.  first and last egg mass detected per year
5.  length of breeding season

```{r statistics_dataframe}
statistics <-data|>
  group_by(BRDYEAR,Watershed,LocationID)|>
  summarize(survey_count = n_distinct(EventGUID),
            mean_egg_num = (sum(NumberofEggMasses, na.rm = TRUE)/survey_count),
            total_egg_num =sum(NumberofEggMasses, na.rm = TRUE),
            firstEgg = min(dayOfWY), 
            lastEgg = max(dayOfWY), 
            breedingLength = max(dayOfWY) - min(dayOfWY))
statistics
```

# plots

#### eggs

```{r mean_NEW_eggs}
#I made a new "data_after2009"data set and plot the trend line using is, instead of the full "data". (since 2000-2005 all have 0 egg masses, and there are no records from 2009 to 2009)
data_after2009 <- data|> 
  filter(BRDYEAR>2009)
data_after2009

statistics_after2009 <- statistics |>
  filter(BRDYEAR>2009)
statistics_after2009
```

```{r mean_NEW_eggs}
#plots of mean number of NEW egg masses per survey
#This code has been updated. If anyone can help adding label info (the mean value) besides each point that would be great. 
ggplot(data = data_after2009,mapping = aes(x = BRDYEAR, y = NumberofEggMasses))+ 
  geom_jitter(alpha=0.1,width = 0.1)+
  stat_summary(fun = "mean",geom = "point",color = "orange", size= 2)+
  geom_smooth(method = "lm", color = "blue",se = FALSE)+
  labs(x = "Year", y = "mean number of new egg masses per survey") 

### above plot facet by watershed
ggplot(data = data_after2009)+ 
  stat_summary(mapping = aes(x = BRDYEAR, y = NumberofEggMasses),
               fun = "mean",geom = "point",color = "black", size= 2)+
  geom_smooth(mapping = aes(x = BRDYEAR, y = NumberofEggMasses), method = "lm", color = "blue",se = FALSE)+
  facet_wrap(~Watershed)+ scale_y_continuous(limits = c(0,1.5))

### above plot facet by site
ggplot(data = data_after2009)+ 
  stat_summary(mapping = aes(x = BRDYEAR, y = NumberofEggMasses),
               fun = "mean",geom = "point",color = "black", size= 2)+
  geom_smooth(mapping = aes(x = BRDYEAR, y = NumberofEggMasses), method = "lm", color = "blue",se = FALSE)+
  facet_wrap(~LocationID)+ scale_y_continuous(limits = c(0,1.5))

```

```{r mean_NEW_eggs}
#plot of total egg masses over time per Watershed
ggplot(data = data_after2009)+ 
  stat_summary(mapping = aes(x = BRDYEAR, y = NumberofEggMasses),
               fun = "sum",geom = "point",alpha = 1/2, color = "black", size= 2) + facet_wrap(~Watershed)+
  labs(x = "Year", y = "total new egg masses")

#plot of total egg masses over time per site
ggplot(data = data_after2009)+ 
  stat_summary(mapping = aes(x = BRDYEAR, y = NumberofEggMasses),
               fun = "sum",geom = "point",alpha = 1/2, color = "black", size= 2) + facet_wrap(~LocationID)+
  labs(x = "Year", y = "total new egg masses")

## above with size of point = survey_count (per site)
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = total_egg_num)) +
  geom_point(aes(size = survey_count), alpha = 1/3) + facet_wrap(~LocationID) +
  labs(x = "Year", y = "total new egg masses")
  
## above with size of point = survey_count
# Do we need this graph? I don't think there is a relationship between survey count and mean of egg masses, unlike total egg masses. We can keep this for now.
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = mean_egg_num)) +
  geom_point(aes(size = survey_count), alpha = 1/3) + facet_wrap(~LocationID)+
  labs(x = "Year", y = "mean new egg masses")


```




#### breeding season

```{r breeding_length}
# timing (days since october 1) of egg laying by year and watershed

eggTiming <- data %>%
  filter(NumberofEggMasses > 0) %>%
  group_by(Watershed, LocationID, BRDYEAR) %>%
  summarize(firstEgg = min(dayOfWY), 
            lastEgg = max(dayOfWY), 
            breedingLength = max(dayOfWY) - min(dayOfWY))
eggTiming

## plot timing v.1
ggplot(data = eggTiming, aes(x = BRDYEAR, y = breedingLength, color = Watershed)) + geom_line()

## improved plot timing v.2
ggplot(data = eggTiming, aes(x = BRDYEAR, y = breedingLength, color = Watershed)) + geom_point() + geom_smooth(method="loess", se=F)

#above plot but facet by watershed
ggplot(data = eggTiming, aes(x = BRDYEAR, y = breedingLength,color = Watershed)) + geom_point() + geom_smooth(method="loess", se=F)+ facet_wrap(~Watershed)


## TODO: correlate breeding season length with rainfall

```

#### sites

```{r num_sites}
#table showing number of different sites for all watersheds
number_of_sites_within_watershed <- data|>
  group_by(Watershed)|>
  summarize(distinct_count = n_distinct(LocationID))
number_of_sites_within_watershed

#plot of the above table.
ggplot(number_of_sites_within_watershed,aes(x=Watershed,y=distinct_count))+
  geom_point()
```

#### survey counts

```{r survey_counts}
# table: number surveys per year (not relevant)
survey_count_by_year <- data|>
  group_by(BRDYEAR)|>
  summarize(survey_count=n_distinct(EventGUID))
survey_count_by_year

# plot: survey count per year across all watersheds/sites
ggplot(data = survey_count_by_year,aes(x=BRDYEAR,y=survey_count))+geom_point()+geom_smooth()+ scale_y_continuous(limits = c(0,200))

# table: number of surveys at each watershed per year
abundance_counts <- data %>%
  group_by(Watershed, BRDYEAR) %>%
  summarize(survey_count = n_distinct(EventGUID))
abundance_counts
```

```{r survey_counts}
# above table but containing only after year 2009. There is a few survey before that and all in one single year in Watershed: redwood creek.
abundance_counts_after2009 <- data_after2009 %>%
  group_by(Watershed, BRDYEAR) %>%
  summarize(survey_count = n_distinct(EventGUID))
abundance_counts_after2009

# plot: number of surveys per year per Watershed 
survey_graph <- ggplot(abundance_counts,aes(x=BRDYEAR,y=survey_count,colour=Watershed))+geom_point()+ facet_wrap(~Watershed)
survey_graph

# above plot containing only year after 2009.
survey_graph_after2009 <- ggplot(abundance_counts_after2009,aes(x=BRDYEAR,y=survey_count,colour=Watershed))+geom_point()+ facet_wrap(~Watershed)
survey_graph_after2009

```

```{r survey_counts}
#heatmap of survey count per year by watershed
ggplot(abundance_counts, aes(x = BRDYEAR, y = Watershed, fill = survey_count)) +
  geom_tile() +
  scale_fill_gradient(low = "#CBEB81", high = "#0E2D8A") +  # Adjust color gradient
  labs(x = "Breeding Year", y = "Watershed", title = "Abundance Heatmap") +  # Add axis labels and title
  theme_minimal()

#above graph containing only year after 2009.
ggplot(abundance_counts_after2009, aes(x = BRDYEAR, y = Watershed, fill = survey_count)) +
  geom_tile() +
  scale_fill_gradient(low = "#CBEB81", high = "#0E2D8A") +  # Adjust color gradient
  labs(x = "Breeding Year", y = "Watershed", title = "Abundance Heatmap") +  # Add axis labels and title
  theme_minimal()

```

```{r survey_count_vs_new_egg}
#number of survey count for each number of new egg masses of all watersheds by year
ggplot(data = statistics, mapping = aes(x = total_egg_num)) + 
  geom_freqpoly(mapping = aes(colour = BRDYEAR))+
  scale_x_continuous(limits = c(0, 170))
```

##### observers

```{r obsv_total}
# investigating number of observers (obsv_total)
## histogram of obsv_total
ggplot(data = data, aes(x = obsv_total)) + geom_histogram()
```

#### environmental covariates

```{r}
#counting the number of non-NA values for the environmental variables we are looking at
## SALINITY: There are 727 non-NA salinity data. We can adjust based on what Darren replied in email. 
length(na.omit(data$WaterSalinity))

## OPEN WATER： There are 3030 non-NA open water data
length(na.omit(data$PercentOpenWater))

## DEPTH (AVG & MAX): Darren said they are not that relevant; if we use we should use max depth (the one with more data)

#there are 2020 non-zero average depth data
length(na.omit(data$AvgD))

#there are 3099 non-zero maximum depth data
length(na.omit(data$MaxD))

#correlation between average depth and max depth is significantly positive 
regression_depth=lm(AvgD~MaxD,data=new_egg)
summary(regression_depth)

library(ggpubr)
ggscatter(new_egg, x = "MaxD", y = "AvgD",
          add = "reg.line",cor.coef=TRUE,coor.method=" ")
```
