---
title: "2_EDA"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
setwd(here::here("code"))
source("1_data_prep.R")
```

# TODO

-   data = filtered by 2+ surveys, from the main 7 watersheds
-   *delete anything that is filtering/selecting 7 watersheds in this file*

# data manipulation

tables that are helpful for creating many graphs

```{r summary_new_eggs}
#summary information of new egg masses count
new_egg <-data[data$OldMass=="FALSE",]
new_total_egg_masses <- new_egg$NumberofEggMasses

#total number of egg masses is 4000
sum(new_total_egg_masses, na.rm = TRUE)

#maximum number of egg masses for a given survey is 25, while the mean is 0.84
summary(new_total_egg_masses,na.rm = TRUE)
```

#### dataframe: "statistics"

1.  number of survey
2.  mean number of new egg masses per survey
3.  total number of egg masses per watershed per site per year
4.  first and last egg mass detected per year
5.  length of breeding season

```{r statistics_dataframe}
statistics <-new_egg|>
  group_by(BRDYEAR,Watershed,LocationID)|>
  summarize(survey_count = n_distinct(EventGUID),
            mean_egg_num = (sum(NumberofEggMasses, na.rm = TRUE)/survey_count),
            total_egg_num =sum(NumberofEggMasses, na.rm = TRUE),
            firstEgg = min(dayOfWY), 
            lastEgg = max(dayOfWY), 
            breedingLength = max(dayOfWY) - min(dayOfWY)
            )
```

# plots

#### eggs

```{r total_eggs}
# plot total number of eggs per year by site (shortened x-axis for view)
## is this the true sum? double counts?
number_of_eggs_per_year_by_site <-
  ggplot(data = totalEggsPerYear, aes(x = BRDYEAR, y = totalEggs)) +
  geom_point() + facet_wrap(totalEggsPerYear$LocationID) + 
  scale_x_continuous(breaks = seq(min(totalEggsPerYear$BRDYEAR), max(totalEggsPerYear$BRDYEAR), by = 6))
number_of_eggs_per_year_by_site 
```

```{r mean_NEW_eggs}
#plots of mean number of NEW egg masses per survey
ggplot(data = statistics)+ 
  stat_summary(mapping = aes(x = BRDYEAR, y = mean_egg_num),
               fun = "mean",geom = "point",color = "blue", size= 2)+
  labs(x = "Year", y = "mean number of new egg masses per survey")

## TODO: alternate plot of NEW egg masses = all observations (scatter plot) with a TRENDLINE (= mean)
### ALL 7 watersheds

#### REVISE --> graph y= NumberofEggMasses, x= year
ggplot(data = statistics, aes(x = BRDYEAR, y = mean_egg_num)) + geom_point() +
  labs(x = "Year", y = "mean number of new egg masses") + geom_smooth(method="lm")+ggtitle("mean number of new egg masses per survey (all watersheds)")

### by watershed


#plot of total egg masses over time per site
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = total_egg_num)) +
  geom_point(alpha = 1/3) + facet_wrap(~LocationID) +
  labs(x = "Year", y = "total new egg masses")

## above with size of point = survey_count
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = total_egg_num)) +
  geom_point(aes(size = survey_count), alpha = 1/3) + facet_wrap(~LocationID) +
  labs(x = "Year", y = "total new egg masses")
  
#plot of mean egg masses over time per site
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = mean_egg_num)) +
  geom_point(alpha = 1/3) + facet_wrap(~LocationID)+
  labs(x = "Year", y = "mean new egg masses")

## above with size of point = survey_count
ggplot(data = statistics, mapping = aes(x = BRDYEAR, y = mean_egg_num)) +
  geom_point(aes(size = survey_count), alpha = 1/3) + facet_wrap(~LocationID)+
  labs(x = "Year", y = "mean new egg masses")

```

#### breeding season

```{r breeding_length}
# timing (days since october 1) of egg laying by year and watershed
eggTimingNoZero <- eggTiming %>% 
  filter(breedingLength > 0) %>% 
  group_by(Watershed, BRDYEAR) %>% 
  summarize(meanFirstEgg = mean(firstEgg), meanLastEgg = mean(lastEgg), meanLength = mean(lastEgg) - mean(firstEgg))

## TODO: change meanLength by SITE 

## plot timing v.1
ggplot(data = eggTimingNoZero, aes(x = BRDYEAR, y = meanLength, color = Watershed)) + geom_line()

## improved plot timing v.2
ggplot(data = eggTimingNoZero, aes(x = BRDYEAR, y = meanLength, color = Watershed)) + geom_point() + geom_smooth(method="lm", se=F)

## TODO: correlate breeding season length with rainfall
### start of season with rainfall

```

#### sites

```{r num_sites}
#table showing number of different sites for all watersheds
number_of_sites_within_watershed <- data|>
  group_by(Watershed)|>
  summarize(distinct_count = n_distinct(LocationID))
number_of_sites_within_watershed

#plot of the above table.
ggplot(number_of_sites_within_watershed,aes(x=Watershed,y=distinct_count))+
  geom_point()
```

#### survey counts

```{r survey_counts}
# table: number surveys per year (not relevant)
survey_count_by_year <- data|>
  group_by(BRDYEAR)|>
  summarize(survey_count=n_distinct(EventGUID))
survey_count_by_year

# plot: survey count per year across all watersheds/sites
ggplot(data = survey_count_by_year,aes(x=BRDYEAR,y=survey_count))+geom_point()+geom_smooth()

# table: number of surveys at each watershed per year
abundance_counts <- data %>%
  group_by(Watershed, BRDYEAR) %>%
  summarize(survey_count = n_distinct(EventGUID))
abundance_counts

# plot: number of surveys per year at rich watersheds
## only at watersheds with decent amount of data (7 watersheds in total) 
rich_watersheds <-abundance_counts|>
  filter(Watershed == "Kanoff Creek" | Watershed == "Laguna Salada" | Watershed =="Milagra Creek"|
           Watershed == "Redwood Creek" | Watershed == "Rodeo Lagoon" | Watershed=="Tennessee Valley" |
           Watershed == "Wilkins Gulch")

rich_graph <- ggplot(rich_watersheds,aes(x=BRDYEAR,y=survey_count,colour=Watershed,group=Watershed))+geom_point()
rich_graph <- rich_graph+ facet_wrap(~Watershed)
rich_graph

#heatmap of survey count per year by watershed
## edit x-axis labels
ggplot(abundance_counts, aes(x = BRDYEAR, y = Watershed, fill = survey_count)) +
  geom_tile() +
  scale_fill_gradient(low = "#CBEB81", high = "#0E2D8A") +  # Adjust color gradient
  labs(x = "Breeding Year", y = "Watershed", title = "Abundance Heatmap") +  # Add axis labels and title
  theme_minimal()
```

```{r survey_count_vs_new_egg}
#number of survey count for each number of new egg masses of all watersheds by year
ggplot(data = statistics, mapping = aes(x = total_egg_num)) + 
  geom_freqpoly(mapping = aes(colour = BRDYEAR))+
  scale_x_continuous(limits = c(0, 170))
```

##### observers

```{r obsv_total}
# investigating number of observers (obsv_total)
## histogram of obsv_total
ggplot(data = data, aes(x = obsv_total)) + geom_histogram()
```

#### environmental covariates

```{r}
#counting the number of non-NA values for the environmental variables we are looking at 
## SALINITY
data$WaterSalinity
non_na_Salinity <- na.omit(data$WaterSalinity)
length(non_na_Salinity)
sum(non_na_Salinity != 0)
#there are 1200 salinity data, only 92 of them is non-zero. 

data |>
  filter(data$WaterSalinity>0)|>
  group_by(Watershed)|>
  summarise(count=n())
#the distribution of these 92 non-zero salinity data by watershed.

## OPEN WATER
data$PercentOpenWater
non_na_open <- na.omit(data$PercentOpenWater)
length(non_na_open)
sum(non_na_open != 0)
#there are 4937 non-zero open water data

## DEPTH (AVG & MAX)
data$AvgD
non_na_depth <- na.omit(data$AvgD)
length(non_na_depth)
#there are 3404 non-zero open water data

data$MaxD
non_na_depth2 <- na.omit(data$MaxD)
length(non_na_depth2)
#there are 5091 non-zero open water data

#correlation between average depth and max depth is significantly positive 
regression_depth=lm(AvgD~MaxD,data=new_egg)
summary(regression_depth)

ggscatter(new_egg, x = "MaxD", y = "AvgD",
          add = "reg.line",cor.coef=TRUE,coor.method=" ")
```

```{r}
#number of survey count for each number of new egg masses of all watersheds by year
ggplot(data = statistics, mapping = aes(x = total_egg_num)) + 
  geom_freqpoly(mapping = aes(colour = BRDYEAR))+
  scale_x_continuous(limits = c(0, 170))
```
